Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2021-09-08 hpp.tex
Finish applying corrections.

Today, robots in industrial manufacturing are mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. However, automating a task with some variability is very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step to the simplification of industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or the presence of unknown
obstacles. The work only covers the motion planning -- and
more accurately the manipulation planning -- issue. The integration in a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modelling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found by following this \href{https://humanoid-path-planner.github.io/hpp-doc}{link}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

@@

Motion planning has given rise to a lot of research work for the past decades.
The problem consists in finding a collision-free path for a given system in
an environment populated with obstacles. The field covers a large variety of
different applications ranging from navigation for autonomous vehicles in
partially known environments~\cite{DarpaUrbanChallenge} to path planning for
deformable objects~\cite{LamKav2001}, \cite{RouFerTai2020} and many other applications like coverage path planning~\cite{coverage2001}, \cite{coverage2013}, or pursuit evasion planning~\cite{pursuit-evasion1999}.

Planning motions for high dimensional robots like humanoid robots or multi-arm
systems has been proven to be of a very high complexity~\cite{SchSha1983}, \cite{Canny1983}. Starting in the 1990's random sampling methods have been proposed
to solve the problem, trading the completeness property against efficiency in
solving problems in high dimensional configuration spaces \cite{KSLO1996}, \cite{HsuLatMot99}, \cite{KufLav00}. The latter methods are said probabilistically complete since the probability to find a solution if one exists converges to 1 when the time of computation tends to infinity. Since then, asymptotically optimal random sampling algorithms have
been proposed~\cite{KarFra2011}.

@@

Some systems are subject to nonlinear constraints. These nonlinear
constraints define sub-manifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping a same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another way to sample
configurations on a constrained manifold consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this is possible. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference resides in the fact that the nodes built on the
tangent space are not immediately projected on the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} propose a general framework to plan task constrained
motions in the presence of moving obstacle. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

Manipulation planning is a particular instance of path planning where
some objects are moved by some robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
nonlinear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

The first attempt to solve manipulation planning problems using random sampling
was proposed by {\color{blue}Siméon \textit{et al}}~\cite{simeon2004ijrr} where a reduction property reduces the
complexity of the problem.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, that is
a solution where each object is grasped at the most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

Our work shares many ideas with {\color{blue}Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to the latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We address indeed only prehensile manipulation.

@@

Open-source software platforms are an important tool to make possible fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. The most popular one is undoubtedly OMPL \cite{ompl2012}. OMPL integrates many randomized path planning algorithms and is widely used for teaching. Recently, {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to nonlinear constraints.

@@

One of the main differences between our solution and the previously cited ones resides in the way manipulation constraints are compiled into a graph. To our knowledge none of the previous solutions can handle a variety of problems as large as those described in Section~\ref{sec:benchmarks}.

@@

This representation although well suited for fixed base manipulator
arms is not well adapted for robots with a mobile base like wheeled
mobile robots, aerial robots or legged robots, since the mobility of the base
cannot be correctly represented by translation or rotation joints.
Representing a free-flying object by three virtual translations followed by
three virtual rotations called roll, pitch and yaw is indeed a poor workaround
due to the presence of singularities. A good illustration of this is the gimball
lock issue that arose during Apollo~13 flight. To avoid singularities, we
propose the following definition.

\subsubsection{Kinematic chain}

A \textit{kinematic chain} is a tree of \textit{joints} where each
joint represents a mobility of a rigid-body link with respect to
another link or with respect to the world reference frame. To each joint is
associated a configuration space called the \textit{joint configuration space}.
The most common joints with their respective configuration spaces are
\begin{itemize}
\item linear \textit{translation} with configuration space $\reals$,
\item bounded \textit{rotation} with configuration space  $\reals$,
\item unbounded \textit{rotation} with configuration space  $SO(2)$,
\item \textit{planar} joint with configuration space $SE(2)$,
\item \textit{freeflyer} joint with configuration space $SE(3)$.
\end{itemize}
$SO(n)$ and $SE(n)$ respectively stand for \textit{special orthogonal group} and \textit{special Euclidean group}. They represent the group of rotations and the group of rigid-body transformations in $\reals^n$.

\subsubsection{Lie groups}

@@

\paragraph{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and an angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to screw motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

\paragraph{Vector representations}
Each Lie group element is represented by a vector. Rotations are represented by
unit quaternions.

Therefore elements of $SE(3)$ are represented by a vector in $\reals^7$ where the first 3 components represent the image of the origin (vector $\mathbf{t}$ in Equation~\ref{eq:homogeneous matrix}), the last 4 components $(x,y,z,w)$ represent unit quaternion $w + x {i} + y {j} + z {k}$.

@@

\paragraph{Exponential map}

@@

\paragraph{Sum and difference notations}

@@

\subsubsection{Robot configuration space}

@@

\paragraph{Jacobian}

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a {\color{blue} piece-wise} differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity at
  the same time along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the sub-manifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\color{blue}\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\color{blue}\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{He has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ exactly solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, where robots manipulate objects, once an object is grasped, the position of the object can be computed explicitly from the configuration of the robot. In this case, some configuration variables of the system depend on other:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf_{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.

More generally, let us denote by
\begin {itemize}
\item $I_{nq}$ the set of positive integers not greater than $nq=\dim\CS$,
\item $I$ a subset of $I_{nq}$,
\item $\bar{I}$ the complement in $I_{nq}$ of $I$,
\item $|I|$ the cardinal of $I$.
\end {itemize}
If $\conf\in\CS$ is a configuration, we denote by $\conf_{I}\in\reals^{|I|}$ the vector composed of the components of $\conf$ of increasing indices in $I$.
\paragraph {Example} if $\conf = (q_1,q_2,q_3,q_4,q_5,q_6,q_7)$ and $I=\{1,2,6\}$, then $\conf_I=(q_1,q_2,q_6)$, $\conf_{\bar{I}}=(q_3,q_4,q_5,q_7)$.

@@

\paragraph{Example}
If $m=3$, $n=4$, $M=\{2,3\}$ and $N=\{1,2,4\}$,
$$
J = \left(\begin{array}{cccc}
  J_{1,1} &   J_{1,2} &   J_{1,3} &   J_{1,4}\\
  J_{2,1} &   J_{2,2} &   J_{2,3} &   J_{2,4}\\
  J_{3,1} &   J_{3,2} &   J_{3,3} &   J_{3,4}\\
  J_{4,1} &   J_{4,2} &   J_{4,3} &   J_{4,4}
\end{array}\right)
$$
then
$$
J_{M\times N} = \left(\begin{array}{ccc}
  J_{2,1} &   J_{2,2} &   J_{2,4}\\
  J_{3,1} &   J_{3,2} &   J_{3,4}
\end{array}\right)
$$

@@

Let us now detail the iterative computation of~(\ref{eq:jacobian-product}). Let $J$ be the product of of $J_j$ for $j$ from $nc$ to $i+1$.
Note that if $J_i$ and $J$ are square matrices of size $nv$, of the form~(\ref{eq:jacobian-explicit}), $J_i.J$ can be computed by block as follows:
\begin{eqnarray*}
  (J_{i}.J)_{in_{i}\times I_{nv}} &=& J_{in_{i}\times I_{nv}}\\
  (J_{i}.J)_{out_{i}\times I_{nv}} &=& \frac{\partial f_i}{\partial\conf_{in_{i}}}.J_{in_{i}\times I_{nv}}\\
\end{eqnarray*}
and as columns $out$ of $J$ are equal to $0$, left multiplying $J$ by $J_i$ consists in modifying only the following block of $J$:
$$
  (J_{i}.J)_{out_{i}\times in} = \frac{\partial f_i}{\partial\conf_{in_{i}}}.J_{in_{i}\times {in}}\\
$$
Other coefficients of $J_{i}J$ are equal to the corresponding coefficients of $J$.
An implementation of the above Jacobian product can be found \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/e21490c8c713949bd3038dccb6fe02cf254a615f/src/explicit-constraint-set.cc\#L267}{here}.

@@

\paragraph{Important remark} As mentioned in Table~\ref{tab:vector representation}, the configuration and velocity vectors may have different sizes. As a consequence, index sets $in$ and $out$ in Definition~\ref{def:Explicit} correspond to configuration vector indices, while in Expression~(\ref{eq:jacobian-explicit}), they correspond to velocity vector indices. To keep notation simple, we use the same notation for different sets.

@@

Now that we are able to project configurations onto sub-manifolds defined by numerical constraints -- up to some numerical threshold, we need now to define paths on such sub-manifolds. The usual way of doing so is to discretize the path and to project each sample configuration. The shortcoming of this method is that it requires to choose a discretization step at path construction and to lose the continuous information of the path.

@@

When an object is not grasped, it should lie in a stable pose. There are two simple methods to enforce that:
\begin{enumerate}
\item defining virtual grippers in the environment and virtual handles on the object, implicitly defines a discrete set of poses,
\item defining a virtual gripper on an horizontal plane and a virtual handle on the object, and using a grasp with flags $(\false,\false,\true,\true,\true,\false)$ constrains the object to move on a infinite horizontal plane.
\end{enumerate}

@@

\subsection{Automatic construction}

>>>>

Today, robots in industrial manufacturing are mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. However, automating a task with some variability is very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step towards simplifying industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or unknown
obstacles. The work only covers motion planning -- and
more accurately manipulation planning. The integration into a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that can handle implicit and explicit constraints,
\item a manipulation planning algorithm that tackles a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found at \href{https://humanoid-path-planner.github.io/hpp-doc}{https://humanoid-path-planner.github.io/hpp-doc}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

@@

Motion planning has given rise to a lot of research work over the past decades.
The problem consists in finding a collision-free path for a given system in
an environment populated with obstacles. The field covers a large variety of
different applications ranging from navigation for autonomous vehicles in
partially known environments~\cite{DarpaUrbanChallenge} to path planning for
deformable objects~\cite{LamKav2001}, \cite{RouFerTai2020} and many other applications like coverage path planning~\cite{coverage2001}, \cite{coverage2013}, or pursuit evasion planning~\cite{pursuit-evasion1999}.

Planning motions for high dimensional robots like humanoid robots or multi-arm
systems has been shown to be highly complex~\cite{SchSha1983}, \cite{Canny1983}. Starting in the 1990's random sampling methods have been proposed
to solve the problem, trading the completeness property against efficiency in
solving problems in high dimensional configuration spaces \cite{KSLO1996}, \cite{HsuLatMot99}, \cite{KufLav00}. The latter methods are said to be probabilistically complete since the probability to find a solution if one exists converges to 1 when the time of computation tends to infinity. Since then, asymptotically optimal random sampling algorithms have
been proposed~\cite{KarFra2011}.

@@

Some systems are subject to nonlinear constraints. These nonlinear
constraints define sub-manifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping the same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another method consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this can be done. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference lies in the fact that the nodes built on the
tangent space are not immediately projected onto the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} put forward a general framework to plan task-constrained
motions in the presence of moving obstacles. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

Manipulation planning is a particular instance of path planning where
some objects are moved by robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is subject to
nonlinear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

The first attempt to solve manipulation planning problems using random sampling
was proposed by {\color{blue}Siméon \textit{et al}}~\cite{simeon2004ijrr} where a reduction property simplifies the problem.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, where each object is grasped at the most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

Our work shares many ideas with {\color{blue}Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to the latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We only address prehensile manipulation.

@@

Open-source software platforms are an important tool to enable fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. Undoubtedly the most popular one is OMPL \cite{ompl2012} which integrates many randomized path planning algorithms and is widely used for teaching purposes. Recently, {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to nonlinear constraints.

@@

One of the main differences between our solution and the previously cited ones lies in the way manipulation constraints are compiled into a graph. To our knowledge, none of the previous solutions can handle such a variety of problems as large as those described in Section~\ref{sec:benchmarks}.

@@

Although well-suited for fixed base manipulator
arms, this representation is ill-suited for robots with a mobile base like wheeled
mobile, aerial or legged robots, since the mobility of the base
cannot be correctly represented by translation or rotation joints.
Representing a free-flying object by three virtual translations followed by
three virtual rotations referred to as roll, pitch and yaw is indeed a poor workaround
due to the presence of singularities. A good illustration of this is the gimball
lock issue that arose during Apollo~13 flight. To avoid singularities, the following definition is proposed.

\subsection{Kinematic chain}

A \textit{kinematic chain} is a tree of \textit{joints} where each
joint represents the mobility of a rigid-body link with respect to
another link or to the world reference frame.
A configuration space called the \textit{joint configuration space} is
associated to each joint.
The most common joints with their respective configuration spaces are
\begin{itemize}
\item linear \textit{translation} with configuration space $\reals$,
\item bounded \textit{rotation} with configuration space  $\reals$,
\item unbounded \textit{rotation} with configuration space  $SO(2)$,
\item \textit{planar} joint with configuration space $SE(2)$,
\item \textit{freeflyer} joint with configuration space $SE(3)$.
\end{itemize}
$SO(n)$ and $SE(n)$ respectively stand for \textit{special orthogonal group} and \textit{special Euclidean group}. They represent the group of rotations and the group of rigid-body transformations in $\reals^n$.

\subsection{Lie groups}

@@

\subsubsection{Geometric interpretations}
\begin{itemize}
\item $\reals$ (and by trivial generalization $\reals^n$): the neutral element is $0$. The tangent space at 0 is isomorphic to $\reals$ and
  $$
  \forall \theta\in\reals,\  \exp(\theta) = \theta.
  $$
\item $SE(3)$: an element $g$ of $SE(3)$ can be seen as the position of a moving frame in a fixed reference frame. A point $\mathbf{x}\in\reals^3$ is mapped to $g(\mathbf{x})$. Note that $\mathbf{x}$ is also the coordinate vector of $g(\mathbf{x})$ in the moving frame $g$. If $\mathbf{v}$, $\omega$ are a linear and angular velocities at the origin, $(\mathbf{v},\omega)$ is \textit{transported} to $g$
  as the same linear and angular velocities expressed in the moving frame. In other words, if
  \begin{equation}\label{eq:homogeneous matrix}
  M=\left(\begin{array}{ll} R & \mathbf{t}\\ 0 & 1\end{array}\right)
    \end{equation}
  with $R\in SO(3)$ and $\mathbf{t}\in\reals^3$  is the homogeneous matrix
  representing $g$, and $(\mathbf{v},\mathbf{\omega})$ is a velocity in $T_{I_3}SE(3)$, the velocity \textit{transported} to $g$ corresponds to linear and angular velocities $R\mathbf{v}$ and $R\omega$ of the moving frame.
  Integral curves of the vector field mentioned in item 2 above correspond to screw motions of constant velocity expressed in the moving frame.
\end{itemize}
$SE(2)$, $SO(3)$, and $SO(2)$ are (or are isomorphic to) subgroups of $SE(3)$ and follow the same geometrical interpretation.

\subsubsection{Vector representations}
Each Lie group element is represented by a vector. Rotations are represented by
unit quaternions.

Therefore elements of $SE(3)$ are represented by a vector in $\reals^7$ where the first three components represent the image of the origin (vector $\mathbf{t}$ in Equation~\ref{eq:homogeneous matrix}), the last four components $(x,y,z,w)$ represent unit quaternion $w + x {i} + y {j} + z {k}$.

@@

\subsubsection{Exponential map}

@@

\subsubsection{Sum and difference notations}

@@

\subsection{Robot configuration space}

@@

\subsubsection{Jacobian}

In this \paper, we will make use of the term Jacobian in a generalized way.
If $h$ is a {\color{blue} piece-wise} differentiable function from a Lie group $\liegroup_1$ to a Lie group $\liegroup_2$, and $\conf_1$ an element of $\liegroup_1$, we will denote by $\frac{\partial h}{\partial\conf}(\conf_1)$ the operator that maps velocities in $T_{\conf_1}\liegroup_1$ to the velocity in $T_{h(\conf_1)}\liegroup_2$ \textit{transported by} $h$\footnote{If $\dconf\in T_{\conf_1}\liegroup_1$ is a velocity along a time parameterized curve $\gamma$, $\frac{\partial h}{\partial\conf}(\conf_1)\dconf$ is the velocity along Curve $h(\gamma)$.}.

@@

It is sometimes useful to produce a configuration $\conf$ that satisfies a constraint (or a set of constraints) of type~(\ref{eq:constraint}) from a configuration $\conf_0$ that does not. This action is called the \textit{projection of $\conf_0$ onto the sub-manifold defined by the constraint} and is performed by a Gauss-Newton solver~\cite[Chapter~10]{NoceWrig06} that iteratively linearizes the constraint as follows:
$$
h(\conf_{i+1}) \approx h(\conf_i) + \frac{\partial h}{\partial \conf}(\conf_{i})(\conf_{i+1}{\color{blue}\ominus}\conf_i) = 0
$$
Iterate $\conf_{i+1}$ is computed as follows:
\begin{equation}\label{eq:HierarchicalIterative}
\conf_{i+1} = \conf_i{\color{blue}\ominus}\alpha_i\frac{\partial h}{\partial \conf}^{+}(\conf_{i}) h(\conf_i)
\end{equation}
where $.^{+}$ denotes the Moore Penrose\footnote{who has just been awarded the Nobel Prize.} pseudo inverse, and $\alpha_i$ is a positive real number called the step size. Taking $\alpha_i=1$ solves the linear approximation, but it may not be the best choice in general.

@@

In manipulation planning applications, in which robots manipulate objects, once an object is grasped, the position of the object can be explicitly computed from the configuration of the robot. In this case, some configuration variables of the system depend on other configuration variables:
$$
\conf=(\conf_{rob},\conf_{obj})\in\CS,\ \ \ \ \conf_{obj} = g_{grasp}(\conf_{rob}).
$$
Although this constraint may fit definition~(\ref{eq:constraint}) by defining
\begin{equation}\label{eq:explicit-grasp}
h(\conf) \triangleq \conf_{obj} \ominus g_{grasp}(\conf_{rob}),
\end{equation}
solving this constraint possibly with other constraints using an iterative scheme~(\ref{eq:HierarchicalIterative}) is obviously sub-optimal.

More generally, let us denote by
\begin {itemize}
\item $I_{nq}$ the set of positive integers not greater than $nq=\dim\CS$,
\item $I$ a subset of $I_{nq}$,
\item $\bar{I}$ the complement in $I_{nq}$ of $I$,
\item $|I|$ the cardinal of $I$.
\end {itemize}
If $\conf\in\CS$ is a configuration, we denote by $\conf_{I}\in\reals^{|I|}$ the vector composed of the components of $\conf$ of increasing indices in $I$.
\subsubsection {Example} if $\conf = (q_1,q_2,q_3,q_4,q_5,q_6,q_7)$ and $I=\{1,2,6\}$, then $\conf_I=(q_1,q_2,q_6)$, $\conf_{\bar{I}}=(q_3,q_4,q_5,q_7)$.

@@

\subsubsection{Example}
If $m=3$, $n=4$, $M=\{2,3\}$ and $N=\{1,2,4\}$,
$$
J = \left(\begin{array}{cccc}
  J_{1,1} &   J_{1,2} &   J_{1,3} &   J_{1,4}\\
  J_{2,1} &   J_{2,2} &   J_{2,3} &   J_{2,4}\\
  J_{3,1} &   J_{3,2} &   J_{3,3} &   J_{3,4}\\
  J_{4,1} &   J_{4,2} &   J_{4,3} &   J_{4,4}
\end{array}\right)
$$
then
$$
J_{M\times N} = \left(\begin{array}{ccc}
  J_{2,1} &   J_{2,2} &   J_{2,4}\\
  J_{3,1} &   J_{3,2} &   J_{3,4}
\end{array}\right)
$$

@@

Let us now detail the iterative computation of~(\ref{eq:jacobian-product}). Let $J$ be the product of of $J_j$ for $j$ from $nc$ to $i+1$.
Note that if $J_i$ and $J$ are square matrices of size $nv$, of the form~(\ref{eq:jacobian-explicit}), $J_i.J$ can be computed by block as follows:
\begin{eqnarray*}
  (J_{i}.J)_{in_{i}\times I_{nv}} &=& J_{in_{i}\times I_{nv}}\\
  (J_{i}.J)_{out_{i}\times I_{nv}} &=& \frac{\partial f_i}{\partial\conf_{in_{i}}}.J_{in_{i}\times I_{nv}}\\
\end{eqnarray*}
and as columns $out$ of $J$ are equal to $0$, left multiplying $J$ by $J_i$ consists in modifying only the following block of $J$:
$$
  (J_{i}.J)_{out_{i}\times in} = \frac{\partial f_i}{\partial\conf_{in_{i}}}.J_{in_{i}\times {in}}\\
$$
Other coefficients of $J_{i}J$ are equal to the corresponding coefficients of $J$.
An implementation of the aforementioned Jacobian product can be found \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/e21490c8c713949bd3038dccb6fe02cf254a615f/src/explicit-constraint-set.cc\#L267}{here}.

@@

\subsubsection{Important remark} As mentioned in Table~\ref{tab:vector representation}, the configuration and velocity vectors may have different sizes. As a consequence, index sets $in$ and $out$ in Definition~\ref{def:Explicit} correspond to configuration vector indices, while in Expression~(\ref{eq:jacobian-explicit}), they correspond to velocity vector indices. To keep notation simple, we use the same notation for different sets.

@@

Now that we are able to project configurations onto sub-manifolds defined by numerical constraints -- up to some numerical threshold, we need to define paths on such sub-manifolds. The usual way of doing so is by discretizing the path and projecting each sample configuration. The shortcoming is that it requires choosing a discretization step at path construction thus losing the continuous information of the path.

@@

When an object is not grasped, it should lie in a stable pose. There are two simple methods to enforce that:
\begin{enumerate}
\item defining virtual grippers in the environment and virtual handles on the object, implicitly defines a discrete set of poses,
\item defining a virtual gripper on a horizontal plane and a virtual handle on the object, and using a grasp with flags $(\false,\false,\true,\true,\true,\false)$ constrains the object to move on an infinite horizontal plane.
\end{enumerate}

@@

\subsubsection{Automatic construction}
